---
title: "PA1_template.Rmd"
author: "VibhorVyas"
date: "October 31, 2017"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r}
library(ggplot2)
library(plyr)
activity <- read.csv("activity.csv")
```
##Processing the Data

```{r}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
```
##pulling data without nas
```{r}
clean <- activity[!is.na(activity$steps),]
```
## What is mean total number of steps taken per day?

## summarizing total steps per date
```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```
## Creating the historgram of total steps per day
```{r}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```
## Mean of Steps
```{r}
as.integer(mean(sumTable$Steps))
```
## Median of Steps
```{r}
as.integer(median(sumTable$Steps))
```
## What is the average daily activity pattern?
```{r}
library(plyr)
library(ggplot2)
```
##pulling data without nas
```{r}
clean <- activity[!is.na(activity$steps),]
```
##create average number of steps per interval
```{r}
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))
```
##Create line plot of average number of steps per interval
```{r}
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```
##Maximum steps by interval
```{r}
maxSteps <- max(intervalTable$Avg)
```
##Which interval contains the maximum average number of steps
```{r}
intervalTable[intervalTable$Avg==maxSteps,1]
```
## Imputing missing values

##Number of NAs in original data set
```{r}
nrow(activity[is.na(activity$steps),])
```
## Create the average number of steps per weekday and interval
```{r}
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
```
## Create dataset with all NAs for substitution
```{r}
nadata<- activity[is.na(activity$steps),]
```
## Merge NA data with average weekday interval for substitution
```{r}
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```
## Reorder the new substituded data in the same format as clean data set
```{r}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
```
##Merge the NA averages and non NA data together
```{r}
mergeData <- rbind(clean, newdata2)
```
##Create sum of steps per date to compare with step 1
```{r}
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")
```
## Mean of Steps with NA data taken care of
```{r}
as.integer(mean(sumTable2$Steps))
```
## Median of Steps with NA data taken care of
```{r}
as.integer(median(sumTable2$Steps))
```
## Creating the histogram of total steps per day, categorized by data set to show impact
```{r}
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Red", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "red") )
```
## Are there differences in activity patterns between weekdays and weekends?

## Create new category based on the days of the week
```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
library(lattice) 
## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```
